Foreground Signatures and Mitigation in Neutral Hydrogen Power Spectra From The Reionization Epoch

### 2014-10-10 foreground_delay_spectrum_modeling.tex
From v0.7-MWA-EoR: =================
(*) Incorporated most of Steven Tingay's comments on the first review of draft

%% \title{Detecting Epoch of Reionization in redshifted 21~cm: contamination in EoR window}
%% \title{Limits on the detection of the Epoch of Reionization from MWA observations of the redshifted 21~cm line}
\title{Foreground Signatures and Mitigation in Neutral Hydrogen Power Spectra From Reionization Epoch}

@@

At the end of the recombination epoch, the universe turned completely neutral. This period, referred to as the {\it Dark Ages} in the universe's history, is characterized by collapse of matter under the influence of gravity. And it ended with the formation of the first stars and galaxies which started emitting ultra--violet and X--ray radiation, thereby reionizing the neutral medium in their surroundings. This commenced the epoch of reionization (EoR) --- a period of non--linear growth of matter density perturbations and astrophysical evolution. Studying the epoch of reionization holds the key to understanding this evolution.

Observation of redshifted 21~cm spin flip transition of H{\sc i} has been identified as a direct probe of this epoch \citep{sun72,sco90,mad97,toz00,ili02}. Detecting this signal has recently emerged as a very promising experiment to fill the gaps in our understanding of the universe's history.

Sensitive instruments such as the Square Kilometer Array (SKA) are required for direct observation and tomography of redshifted H{\sc i}. Numerous precursors to the SKA such as the Murchison Widefield Array \citep[MWA;][]{tin13,bow13,lon09}, the Low Frequency Array \citep[LOFAR;][]{van13}, and the Precision Array for Probing the Epoch of Reionization \citep[PAPER;][]{par10} have become operational with enough sensitivity for a statistical detection of the EoR H{\sc i} power spectrum \citep{thy13,bea13,bow06,pob14}.

A key challenge in the statistical detection of power spectrum of this redshifted H{\sc i} 21~cm signal arises from the contamination by Galactic and extragalactic foregrounds \citep[see, e.g.,][]{dim02,zal04,fur06}. \citet{mor04} show that the inherent isotropy and symmetry of the EoR signal in frequency and spatial wavenumber ($k$) space make it distinguishable from sources of contamination which are isolated to certain $k$ modes by virtue of their inherent spectral smoothness \citep{mor06,bow09,liu11,par12,dil13}. Considerable effort has been made towards understanding the $k$--space behavior of foreground signatures in the observed power spectrum \citep{thy13,pob13,mor12,tro12,liu11,dat10,bow09,liu09}. A general model of scattering of smooth spectrum foregrounds to higher $k$--modes by the wide-field response of the instrument, into the so-called `wedge', has emerged that provides a high level explanation for the observed foreground power spectrum. The conservative foreground strategy that has developed alongside this work is to discard $k$--modes which could be contaminated ({\it avoidance}). The more aggressive alternative is to subtract a sky model and regain access to modes lost by {\it avoidance}. In both cases, which parts of the sky are most critical to either {\it avoid} or {\it subtract} has remained largely uncertain. Here, we will focus primarily on extending the {\it avoidance} strategy by identifying foreground components most likely to ``leak'' from foreground modes to EoR modes and demonstrating a scheme for down--weighting these components.

Foregrounds contaminate the EoR H{\sc i} power spectrum by either having intrinsic spectral variability, high instrumental chromaticity, or through approximations in power spectrum analysis. Here we use existing catalogs and a high fidelity instrumental model to capture both foreground and instrumental chromaticity. To decouple our findings from possible analysis effects, such as those pointed out by \citet{haz13}, we compute power spectra using the individual baseline--based approach of \citet{pob13,par14}. This approximates the power spectrum as the inverse Fourier transform of the spectra generated by the instrument's correlator.

Matching real observations from the Murchison Widefield Array with instrumental simulations of the entire known sky, we examine the signature of the power spectrum of each baseline. We report two important findings: the foregrounds most likely to obfuscate the redshifted 21~cm power spectrum are not caused by emission in the central field of view, but rather by bright objects from near the horizon; and, diffuse Galactic emission plays a larger role than previously thought. % We find that the foregrounds most likely to obfuscate the redshifted 21~cm power spectrum are not caused by emission in the central field of view, but rather by bright objects from near the horizon. We find that diffuse Galactic emission plays a larger role then previously thought. 
We quantify this contamination by separating the simulations into components based on type of foreground emission. Using these we arrive at a new method for minimizing the contribution of bright foregrounds which uses prior knowledge of the sky to down--weight adversely contaminated baselines.

In \S\ref{sec:delay-spectrum} we introduce the delay spectrum technique, then in  \S\ref{sec:instrument}, we describe the MWA setup, summarize the observing parameters and present the resulting data. Simulations of these observing parameters are described in \S\ref{sec:modeling} and analyzed by sky location, and type of emission in \S\ref{sec:delay-spectrum-analysis}. In \S\ref{sec:fg-grading}, we offer an initial description of a more precise foreground avoidance technique. \S\ref{sec:summary} summarizes our work and findings.

@@

Interferometer array data known as {\it visibilities}, $V_f(\overline{\mathbf{u}},f)$, represent correlations between time--series of electric fields measured by different antenna pairs with separation vectors $\overline{\mathbf{x}}$ and then Fourier transformed along time axis to obtain a spectrum along frequency ($f$) axis. Here, $\overline{\mathbf{u}}\equiv \overline{\mathbf{x}}f/c \equiv \overline{\mathbf{x}}/\lambda$, $c$ is the speed of light and $\lambda$ is the wavelength. $\overline{\mathbf{u}}$ is related to transverse spatial frequency modes ($\overline{\mathbf{k}}_\perp$) of brightness distribution of the sky as $\overline{\mathbf{k}}_\perp \equiv 2\pi\overline{\mathbf{u}}/D(z)$, where $D(z)$ is the transverse comoving distance at a redshift $z$. If $I(\overline{\mathbf{l}},f)$ is the emission at different frequencies on the sky as a function of direction unit vector ($\overline{\mathbf{l}}$) specified as direction cosines, $V_f(\overline{\mathbf{u}},f)$ represents the Fourier decomposition of $I(\overline{\mathbf{l}},f)$ attenuated by the antenna's angular power--pattern $A(\overline{\mathbf{l}},f)$, into transverse spatial frequency modes. Taking into account the instrumental bandpass weights $W_f(f)$, this relation can be written as:
\begin{align}\label{eqn:obsvis}
  V_f(\overline{\mathbf{u}},f) &= \iint\limits_\textrm{sky} A(\overline{\mathbf{l}},f)\,I(\overline{\mathbf{l}},f)\,W_f(f)\,e^{-i2\pi \overline{\mathbf{u}}\cdot\overline{\mathbf{l}}}\,\dif^2 \overline{\mathbf{l}}.
\end{align}

We define the {\it delay spectrum} $V_\eta(\overline{\mathbf{u}},\eta)$ to be the inverse Fourier transform of $V_f(\overline{\mathbf{u}},f)$ along the frequency coordinate:
\begin{align}\label{eqn:delay-transform}
  V_\eta(\overline{\mathbf{u}},\eta) &= \int V_f(\overline{\mathbf{u}},f)\,W'_f(f)\,e^{i2\pi\eta f}\,\dif f,
\end{align}
where, $W'_f(f)$ is a frequency window weighting function which can be chosen to control the quality of the delay spectrum \citep{thy13,ved12}. $\eta$ represents the signal delay between antenna pairs:
\begin{equation}\label{eqn:delay}
  \eta = \frac{\overline{\mathbf{x}}\cdot\overline{\mathbf{l}}}{c} = \frac{\overline{\mathbf{u}}\cdot\overline{\mathbf{l}}}{f}.
\end{equation}
Note that as the visibility is a fully complex quantity, its Fourier transform is fully Nyquist sampled over positive and negative delays. Negative and positive delay\footnote{The terms delay and lag are used interchangeably in this paper.} correspond to the two hemispheres of the sky as transected by the antenna spacing vector. In this paper, $\overline{\mathbf{u}}$ and $\overline{\mathbf{x}}$ are assumed to be on a coordinate system aligned with the local east, north (along local meridian) and upward directions at the telescope site. Hence, a perfectly Eastward oriented antenna spacing will observe an object in the Eastern sky at positive delay, and conversely for an object in the Western sky. Similarly, an object in the Northern sky will appear at a positive delay at an antenna spacing oriented Northward.

Since we are studying a redshifted H{\sc i} spectral line, $f$ is a measure of cosmological distance along the line of sight. Consequently, $\eta$ is a measure of line of sight spatial frequency mode given by $k_\parallel\approx 2\pi\eta\,f_{21}H_0\,E(z)/(c(1+z)^2)$, where $f_{21}$ is the rest frame frequency of 21~cm spin flip transition of H{\sc i}, and $H_0$ and $E(z)\equiv [\Omega_\textrm{M}(1+z)^3+\Omega_\textrm{k}(1+z)^2+\Omega_\Lambda]^{1/2}$ are standard terms in cosmology. We use $H_0=100\,h$~km~s$^{-1}$~Mpc$^{-1}$ and assume a flat $\Lambda$CDM universe with $\Omega_\textrm{M}=0.27$ and $\Omega_\Lambda=0.73$.

Thus, $|V_\eta(\overline{\mathbf{u}},\eta)|^2$ is directly related to the spatial power spectrum of EoR H{\sc i} distribution, $|V(\overline{\mathbf{k}}_\perp,k_\parallel)|^2$ \citep{mor04}. However, foreground objects also emit in radio frequencies. Due to their spatial and spectral (chromatic) structure, the measured $V_\eta(\overline{\mathbf{u}},\eta)$ is contaminated by foreground radio objects \citep{thy13,tro12,mor12,bow09}. These contaminations have to be characterized precisely in order to reduce their impact on EoR H{\sc i} power spectrum detection sensitivity.

In summary, $V_\eta(\overline{\mathbf{u}},\eta)$ is closely related to visibilities, $V_f(\overline{\mathbf{u}},f)$, the basic data blocks measured by each antenna pair, as seen from equation~\ref{eqn:delay-transform}. It captures all the effects of EoR H{\sc i} signal corruption caused by foregrounds and the instrument. At the same time, it is directly related to the sought power $|V(\overline{\mathbf{k}}_\perp,k_\parallel)|^2$ in $k$--modes containing critical information about spatial scales. For our analysis, we use an equivalent quantity, $V_\eta(\overline{\mathbf{x}},\eta) \equiv V_\eta(\lambda\overline{\mathbf{u}},\eta)$. For purposes of illustration, we often use a simplified quantity $|V_\eta(|\overline{\mathbf{x}}|,\eta)|$.

Figure~\ref{fig:fourier-space} illustrates the Fourier space in which the delay (and power) spectra of H{\sc i} from the EoR are presented in this paper. $|\overline{\mathbf{x}}|$ and $k_\perp$, denoting spatial scales in the transverse direction, form the $x$--axis. $\eta$ and $k_\parallel$, denoting spatial scales along line of sight form the $y$--axis. Foreground emission maps to a wedge--shaped region in Fourier space, hereafter referred to as the {\it foreground wedge}, whose boundaries are determined by antenna spacing and the light travel time across it. These boundaries are called horizon delay limits and are shown by solid lines. Spectral transfer function of the instrument convolves the {\it foreground wedge} and stretches it further (unshaded narrow strips bounded by solid and dashed lines) along $\eta$--axis. This effect has been quantified in \citet{thy13}. The width of this narrow strip is inversely proportional to operating bandwidth. Region of Fourier space excluding the {\it foreground wedge} and the narrow strips is usually called the {\it EoR window}. These are shown in light and medium shades. However, the instrument used in our study, namely the MWA, has a passband constructed using coarse channels with a characteristic shape giving rise to periodic gratings thereby resulting in repetitions of the {\it foreground wedge} at multiples of 0.78~$\mu$s. The underlying cause of this effect will be discussed later in this paper.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/fourier_space_185.0_MHz_30.7_MHz.eps}
\caption{Illustration of Fourier space in which delay (and power) spectra of H{\sc i} signals from the EoR are presented. The $x$--axis is denoted by $|\overline{\mathbf{x}}|$ (antenna spacing) or $k_\perp$ (transverse wavenumber). The $y$--axis denoted by $\eta$ (delay) or $k_\parallel$ (line--of--sight wavenumber). The dark shaded region is referred to as {\it foreground wedge} where smooth spectrum foregrounds reside. Its boundaries (solid lines), given by light travel time for corresponding antenna spacings, are referred to as horizon delay limits. Regions outside the {\it wedge} and the narrow white unshaded strips (caused by convolution with the instrumental transfer function) are expected to be relatively free of foreground contamination. Due to an undesirable grating pattern (dot--dashed lines) of the instrument used in this study, namely the MWA, we identify a restricted region of high EoR sensitivity (medium shade) and refer to it as the {\it EoR window} in this paper. \label{fig:fourier-space}}
\end{figure}

@@

The MWA was completed in 2012, and after commissioning, began its EoR observing program in 2013. In its final configuration the MWA is a 128--tile interferometer array capable of observing a 30.72~MHz instantaneous band anywhere in the range 80--300~MHz. Each tile is a phased array of 16 dipoles each in the shape of a bow--tie. This yields a primary field of view $\gtrsim$~20\arcdeg wide and multiple secondary lobes for the tile. The tile arrangement is a centrally condensed core of $\sim$~300~m --- there are many spacings in the range 5--50~m --- and a radial density that falls off as the inverse of the radius with the longest baselines at 3~km \citep{bea12}. Here we focus on antenna spacings $|\overline{\mathbf{x}}| \le 200$~m (spatial scales relevant to reionization) and a frequency range 170--200~MHz. The center frequency 185~MHz corresponds to a redshift, $z\simeq 6.68$.

The MWA passband of width 30.72~MHz is divided coarsely into 24 1.28~MHz chunks with each chunk weighted by a digital filter. The coarse channel shape is obtained using a 8--tap polyphase filter bank (PFB) and a Kaiser window with parameter $\beta=5$. After correcting for the shape of these coarse channels, the channels at the edges of these chunks are flagged because they are known to be possibly contaminated by aliasing at a low level. Any spectral flags, or other known weights constitutes $W_f(f)$ in equation~\ref{eqn:obsvis}.

@@

The MWA targets two primary low-foreground fields for reionization observations. Here, we focus on the field at RA = 0h, Dec = $-30$\arcdeg. The MWA tracks a patch of sky through antenna beams formed and steered electronically by controlling delay settings of an array of dipoles in an MWA tile. The pointing system is capable of steering to points on a regular $\sim$7\arcdeg grid. During observations we allow the sky to drift across the nearest available pointing, shifting between grid points as necessary (once in $\sim 30$~minutes). This process is repeated throughout the course of the observation $\approx 4.86$~hours. Figure~\ref{fig:pointings} illustrates how the apparent pointing oscillates around the desired pointing direction.

The observations shown here were observed on Aug 23, 2013. This day has been chosen as a reference data set for comparison within the MWA EoR collaboration and was selected on the basis of low number of known system errors from among a long stretch of stable observing.

Here we have chosen two, 112 second long sections from this night for detailed study. These were chosen to provide a selection of possible foreground and instrumental conditions. As an example of ``nominal'' observing setup we choose a zenith pointing; as an example of poor foreground conditions, we choose a pointing when the field is $\sim 2$~hours from zenith. This pointing has a significantly higher secondary lobe structure and is observed when the bright galactic center is well above the horizon.

The {\it local sidereal time} (LST) at these two snapshots are shown in Figure~\ref{fig:pointings} as vertical lines at $-1.92$~hours (i.e., 22.08~hours) and 0.09~hours which are hereafter denoted as {\it off--zenith} and {\it zenith} pointings respectively. %These pointings (phased array power--patterns) are centered at RA = 4\fdg 387, Dec = -29\fdg 86 and RA = -28\fdg 8, Dec = -26\fdg 701, whereas the visibilities themselves are phased to zenith corresponding to RA = -28\fdg 8, Dec = -26\fdg 701, and RA = 1\fdg 35, Dec = -26\fdg 701 respectively.

@@

The data are flagged\footnote{\url{http://sourceforge.net/p/aoflagger}} for interference \citep{off12,off10}, removing 3\% of the data and averaged in time  and frequency from the raw 0.5~s, 40~kHz to 2s, 80~kHz. These data are then calibrated to a simulation of the sky containing 2420 point--like objects selected from the MWA Commissioning Survey (MWACS; Hurley-Walker et al. 2014, submitted). MWACS was compiled using observations taken with subsets of the array during the final buildout and commissioning. It has a flux density limit of 25~mJy and a declination range of -12\arcdeg to -40\arcdeg evenly covering the field of view of the observations reported here. The objects used in calibration are selected to lie inside the 5\% contour of the primary lobe of the tile power--pattern. The calibration algorithm -- based on a forward modeling software by \citet{sul12} -- computes complex gain solutions per channel per antenna averaged to two minute intervals. The solutions are still fairly low signal--to--noise so we iteratively project along the antenna and frequency dimensions to capture the relatively independent passband and antenna--to--antenna variation. First, we average the channel gains over all antennas to obtain a high signal--to--noise measurement of the bandpass. After applying this single passband, we do a second round of calibration and fit second and first order polynomials for amplitude and phase respectively for each antenna. This flattens any residual variation in bandpass and removes small phase slopes due to variation in cable delay. Finally, we fit for an additional phase known to be caused by reflection in 150~m length cables. %possibly could add elaboration, if I knew exactly which version of the cal was done here. doesn't really matter

@@

We show results of delay spectrum from MWA data after deconvolution along delay axis in Figure~\ref{fig:fhd_data}. The principle feature is the {\it foreground wedge}, whose boundaries, marked by solid white lines, increase linearly with light travel time corresponding to baseline length. Note that the wedge is not filled out uniformly. Each baseline has a different orientation and resulting response to different objects in the sky. It is this variation that we will explore below.

@@

Figure~\ref{fig:noiseless-dsm-delay-spectrum} shows the amplitude of delay spectrum, $|V_\eta(|\overline{\mathbf{x}}|,\eta)|$, simulated with only the diffuse foreground model, omitting compact sources. The top and bottom panels correspond to {\it off--zenith} and {\it zenith} pointings respectively.

>>>>

%% \title{Detecting Epoch of Reionization in redshifted 21~cm: contamination in EoR window}
%% \title{Limits on the detection of the Epoch of Reionization from MWA observations of the redshifted 21~cm line}
\title{Foreground Signatures and Mitigation in Neutral Hydrogen Power Spectra From The Reionization Epoch}

@@

At the end of the recombination epoch, the Universe was completely neutral. This period, referred to as the {\it Dark Ages} in the Universe's history, is characterized by the localized accumulation of matter under the influence of gravity. And it ended with the formation of the first stars and galaxies which started emitting ultra--violet and X--ray radiation, thereby reionizing the neutral medium in their surroundings. This commenced the epoch of reionization (EoR) --- a period of non--linear growth of matter density perturbations and astrophysical evolution. Studying the EoR holds the key to understanding this evolution.

Observations of redshifted 21~cm radiation generated by the spin flip transition of H{\sc i} has been identified as a direct probe of the EoR \citep{sun72,sco90,mad97,toz00,ili02}. Detecting this signal has recently emerged as a very promising experiment to fill the gaps in our understanding of the Universe's history.

Sensitive instruments such as the Square Kilometre Array (SKA) are required for direct observation and tomography of redshifted H{\sc i}. Numerous precursors to the SKA such as the Murchison Widefield Array \citep[MWA;][]{tin13,bow13,lon09}, the Low Frequency Array \citep[LOFAR;][]{van13}, and the Precision Array for Probing the Epoch of Reionization \citep[PAPER;][]{par10} have become operational with enough sensitivity for a statistical detection of the EoR H{\sc i} power spectrum \citep{thy13,bea13,bow06,pob14}.

A key challenge in the statistical detection of the redshifted H{\sc i} 21~cm signal arises from the contamination by Galactic and extragalactic foregrounds \citep[see, e.g.,][]{dim02,zal04,fur06}. \citet{mor04} show that the inherent isotropy and symmetry of the EoR signal in frequency and spatial wavenumber ($k$) space make it distinguishable from sources of contamination which are isolated to certain $k$ modes by virtue of their inherent spectral smoothness \citep{mor06,bow09,liu11,par12,dil13}.

Considerable effort has been made towards understanding the $k$--space behavior of foreground signatures in the observed power spectrum \citep{thy13,pob13,mor12,tro12,liu11,dat10,bow09,liu09}. A general model of scattering of smooth spectrum foregrounds to higher $k$--modes by the wide-field response of the instrument, into the so-called `wedge', has emerged that provides a high level explanation for the observed foreground power spectrum. The conservative foreground strategy that has developed alongside this work is to discard $k$--modes which could be contaminated ({\it avoidance}). The more aggressive alternative is to subtract a sky model and regain access to modes lost by {\it avoidance}. In both cases, which parts of the sky are most critical to either {\it avoid} or {\it subtract} has remained largely uncertain. Here, we will focus primarily on extending the {\it avoidance} strategy by identifying foreground components most likely to ``leak'' from foreground modes to EoR modes and demonstrating a scheme for down--weighting these components.

Foregrounds contaminate the EoR H{\sc i} power spectrum by either having intrinsic spectral variability, high instrumental chromaticity, or through approximations in power spectrum analyses. Here we use existing catalogs and a high fidelity instrumental model to capture both foreground and instrumental chromaticity. To decouple these effects from possible analysis effects, such as those pointed out by \citet{haz13}, we compute power spectra using the individual baseline--based approach of \citet{pob13} and \citet{par14}. This approximates the power spectrum as the inverse Fourier transform of the spectra generated by the instrument's correlator.

Matching real observations from the Murchison Widefield Array with instrumental simulations of the entire known sky, we examine the signature of the power spectrum of each baseline. We report two important findings: the foregrounds most likely to obfuscate the redshifted 21~cm power spectrum are not caused by emission in the central field of view, but rather by bright objects from near the horizon; and, diffuse Galactic emission plays a larger role than previously thought. % We find that the foregrounds most likely to obfuscate the redshifted 21~cm power spectrum are not caused by emission in the central field of view, but rather by bright objects from near the horizon. We find that diffuse Galactic emission plays a larger role then previously thought. 
We quantify this contamination by separating the simulations into components based on type of foreground emission. We then arrive at a new method for minimizing the contribution of bright foregrounds which uses prior knowledge of the sky to down--weight adversely contaminated baselines.

In \S\ref{sec:delay-spectrum} we introduce the delay spectrum technique. In \S\ref{sec:instrument}, we describe the MWA setup, summarize the observing parameters, and present the resulting data. Simulations using these observing parameters are described in \S\ref{sec:modeling} and analyzed by sky location and type of emission in \S\ref{sec:delay-spectrum-analysis}. In \S\ref{sec:fg-grading}, we offer an initial description of a more precise foreground avoidance technique. We present a summary of our work and findings in \S\ref{sec:summary}.

@@

Interferometer array data known as {\it visibilities}, $V_f(\overline{\mathbf{u}},f)$, represent correlations between time--series of electric fields measured by different antenna pairs with separation vectors $\overline{\mathbf{x}}$ and then Fourier transformed along the time axis to obtain a spectrum along the frequency ($f$) axis. Here, $\overline{\mathbf{u}}\equiv \overline{\mathbf{x}}f/c \equiv \overline{\mathbf{x}}/\lambda$, $c$ is the speed of light, and $\lambda$ is the wavelength of radio waves. $\overline{\mathbf{u}}$ is related to the transverse spatial frequency modes ($\overline{\mathbf{k}}_\perp$) of the brightness distribution of the sky as $\overline{\mathbf{k}}_\perp \equiv 2\pi\overline{\mathbf{u}}/D(z)$, where $D(z)$ is the transverse comoving distance at redshift $z$. If $I(\overline{\mathbf{l}},f)$ is the emission at different frequencies on the sky as a function of direction unit vector ($\overline{\mathbf{l}}$) specified as direction cosines, $V_f(\overline{\mathbf{u}},f)$ represents the Fourier decomposition of $I(\overline{\mathbf{l}},f)$ attenuated by the antenna's angular power--pattern, $A(\overline{\mathbf{l}},f)$, into transverse spatial frequency modes. Taking into account the instrumental bandpass weights $W_f(f)$, this relation can be written as:
\begin{align}\label{eqn:obsvis}
  V_f(\overline{\mathbf{u}},f) &= \iint\limits_\textrm{sky} A(\overline{\mathbf{l}},f)\,I(\overline{\mathbf{l}},f)\,W_f(f)\,e^{-i2\pi \overline{\mathbf{u}}\cdot\overline{\mathbf{l}}}\,\dif^2 \overline{\mathbf{l}}.
\end{align}

We define the {\it delay spectrum}, $V_\eta(\overline{\mathbf{u}},\eta)$, to be the inverse Fourier transform of $V_f(\overline{\mathbf{u}},f)$ along the frequency coordinate:
\begin{align}\label{eqn:delay-transform}
  V_\eta(\overline{\mathbf{u}},\eta) &= \int V_f(\overline{\mathbf{u}},f)\,W'_f(f)\,e^{i2\pi\eta f}\,\dif f,
\end{align}
where, $W'_f(f)$ is a frequency window weighting function which can be chosen to control the quality of the delay spectrum \citep{thy13,ved12} and $\eta$ represents the signal delay between antenna pairs:
\begin{equation}\label{eqn:delay}
  \eta = \frac{\overline{\mathbf{x}}\cdot\overline{\mathbf{l}}}{c} = \frac{\overline{\mathbf{u}}\cdot\overline{\mathbf{l}}}{f}.
\end{equation}
Note that as the visibility is a complex quantity, its Fourier transform is Nyquist sampled over positive and negative delays. Negative and positive delay\footnote{The terms delay and lag are used interchangeably in this paper.} correspond to the two hemispheres of the sky as transected by the antenna spacing vector. In this paper, $\overline{\mathbf{u}}$ and $\overline{\mathbf{x}}$ are assumed to be on a coordinate system aligned with the local east, north (along local meridian) and upward directions at the telescope site. Hence, a perfectly eastward oriented antenna spacing will observe an object in the eastern sky at positive delay, and conversely for an object in the western sky. Similarly, an object in the northern sky will appear at a positive delay at an antenna spacing oriented northward.

Since we are studying a redshifted H{\sc i} spectral line, $f$ is a measure of cosmological distance along the line of sight. Consequently, $\eta$ is a measure of line--of--sight spatial frequency mode given by $k_\parallel\approx 2\pi\eta\,f_{21}H_0\,E(z)/(c(1+z)^2)$, where $f_{21}$ is the rest frame frequency of the 21~cm spin flip transition of H{\sc i}, and $H_0$ and $E(z)\equiv [\Omega_\textrm{M}(1+z)^3+\Omega_\textrm{k}(1+z)^2+\Omega_\Lambda]^{1/2}$ are standard terms in cosmology. We use $H_0=100\,h$~km~s$^{-1}$~Mpc$^{-1}$ and assume a flat $\Lambda$CDM universe with $\Omega_\textrm{M}=0.27$ and $\Omega_\Lambda=0.73$.

Thus, $|V_\eta(\overline{\mathbf{u}},\eta)|^2$ is directly related to the spatial power spectrum of EoR H{\sc i} distribution, $|V(\overline{\mathbf{k}}_\perp,k_\parallel)|^2$ \citep{mor04}. However, foreground objects also emit at radio frequencies. Due to their spatial and spectral (chromatic) structure, the measured $V_\eta(\overline{\mathbf{u}},\eta)$ is contaminated by foreground radio objects \citep{thy13,tro12,mor12,bow09}. These contaminations have to be characterized precisely in order to reduce their impact on EoR H{\sc i} power spectrum detection sensitivity.

In summary, $V_\eta(\overline{\mathbf{u}},\eta)$ is closely related to visibilities, $V_f(\overline{\mathbf{u}},f)$, the basic data blocks measured by each antenna pair, as seen from equation~\ref{eqn:delay-transform}. $V_\eta(\overline{\mathbf{u}},\eta)$ captures all the effects of EoR H{\sc i} signal corruption caused by foregrounds and the instrument. At the same time, it is directly related to the sought power $|V(\overline{\mathbf{k}}_\perp,k_\parallel)|^2$ in $k$--modes containing critical information about spatial scales. For our analysis, we use an equivalent quantity, $V_\eta(\overline{\mathbf{x}},\eta) \equiv V_\eta(\lambda\overline{\mathbf{u}},\eta)$. For purposes of illustration, we often use a simplified quantity, $|V_\eta(|\overline{\mathbf{x}}|,\eta)|$.

Figure~\ref{fig:fourier-space} illustrates the Fourier space in which the delay (and power) spectra of H{\sc i} from the EoR are presented in this paper. $|\overline{\mathbf{x}}|$ and $k_\perp$, denoting spatial scales in the transverse direction, form the $x$--axis. $\eta$ and $k_\parallel$, denoting spatial scales along line--of--sight form the $y$--axis. Foreground emission maps to a wedge--shaped region in Fourier space, hereafter referred to as the {\it foreground wedge}, whose boundaries are determined by antenna spacing and the light travel time across it. These boundaries are called horizon delay limits and are shown by solid lines. The spectral transfer function of the instrument convolves the {\it foreground wedge} and stretches it further (unshaded narrow strips bounded by solid and dashed lines) along $\eta$--axis. This effect has been quantified in \citet{thy13}. The width of this narrow strip is inversely proportional to the operating bandwidth. The region of Fourier space excluding the {\it foreground wedge} and the narrow strips is the so--called {\it EoR window}, shown in light and medium shades. However, the instrument used in our study, namely the MWA, has a passband constructed using coarse channels with a characteristic shape giving rise to periodic gratings thereby resulting in repetitions of the {\it foreground wedge} at multiples of 0.78~$\mu$s. The underlying cause of this effect will be discussed later in this paper.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/fourier_space_185.0_MHz_30.7_MHz.eps}
\caption{Illustration of Fourier space in which delay (and power) spectra of H{\sc i} signals from the EoR are presented. The $x$--axis is denoted by $|\overline{\mathbf{x}}|$ (antenna spacing) or $k_\perp$ (transverse wavenumber). The $y$--axis denoted by $\eta$ (delay) or $k_\parallel$ (line--of--sight wavenumber). The dark shaded region is referred to as the {\it foreground wedge} where smooth spectrum foregrounds reside. Its boundaries (solid lines), given by light travel time for corresponding antenna spacings, are referred to as horizon delay limits. Regions outside the {\it wedge} and the narrow white unshaded strips (caused by convolution with the instrumental transfer function) are expected to be relatively free of foreground contamination. Due to an undesirable grating pattern (dot--dashed lines) of the instrument used in this study, namely the MWA, we identify a restricted region of high EoR sensitivity (medium shade) and refer to it as the {\it EoR window} in this paper. \label{fig:fourier-space}}
\end{figure}

@@

The MWA was completed in 2012 and, after commissioning, began its EoR observing program in 2013. In its final configuration the MWA is a 128--tile interferometer array capable of observing a 30.72~MHz instantaneous band anywhere in the range 80--300~MHz. Each tile is a phased array of 16 dipoles, each in the shape of a bow--tie. This yields a primary field of view $\gtrsim$~20\arcdeg wide and multiple secondary lobes. The array is arranged as a centrally condensed core of $\sim$~300~m --- there are many spacings in the range 5--50~m --- and a radial density that falls off as the inverse of the radius, with the longest baselines at 3~km \citep{bea12}. Here we focus on antenna spacings $|\overline{\mathbf{x}}| \le 200$~m (spatial scales relevant to reionization) and a frequency range 170--200~MHz. The center frequency of 185~MHz corresponds to a redshift, $z\simeq 6.68$.

The MWA passband of width 30.72~MHz is divided coarsely into 24$\times$1.28~MHz sub--bands with each sub--band weighted by a digital filter. The coarse channel shape is obtained using a 8--tap polyphase filter bank (PFB) and a Kaiser window with parameter $\beta=5$. After correcting for the shape of these coarse channels, the fine channels (40~kHz width) at the edges of these sub--bands are flagged because they are known to be possibly contaminated by aliasing at a low level. Any spectral flags, or other known weights constitutes $W_f(f)$ in equation~\ref{eqn:obsvis}.

@@

The MWA targets two primary low-foreground fields for reionization observations. Here, we focus on the field at RA = 0h, Dec = $-30$\arcdeg. The MWA tracks a patch of sky through antenna beams formed and steered electronically by controlling delay settings of the dipoles in an MWA tile. The pointing system is capable of steering to points on a regular $\sim$7\arcdeg grid. During observations we allow the sky to drift across the nearest available pointing, shifting between grid points as necessary (once in $\sim 30$~minutes). This process is repeated throughout the course of the observation $\approx 4.86$~hours. Figure~\ref{fig:pointings} illustrates how the apparent pointing oscillates around the desired pointing direction.

The observations shown here were undertaken on Aug 23, 2013. This day has been chosen as a reference data set for comparison within the MWA EoR collaboration and was selected on the basis of a low number of known system errors from among a long stretch of stable observing.

Here we have chosen two, 112 second long sections from this night for detailed study. These were chosen to provide a selection of possible foreground and instrumental conditions. As an example of a nominal observing setup we choose a zenith pointing; as an example of poor foreground conditions, we choose a pointing when the field is $\sim 2$~hours from zenith. This pointing has a significantly higher secondary lobe structure and is observed when the bright galactic center is well above the horizon.

The {\it local sidereal time} (LST) at these two snapshots are shown in Figure~\ref{fig:pointings} as vertical lines at $-1.92$~hours (i.e., 22.08~hours) and 0.09~hours which are hereafter denoted as {\it off--zenith} and {\it zenith} pointings, respectively. %These pointings (phased array power--patterns) are centered at RA = 4\fdg 387, Dec = -29\fdg 86 and RA = -28\fdg 8, Dec = -26\fdg 701, whereas the visibilities themselves are phased to zenith corresponding to RA = -28\fdg 8, Dec = -26\fdg 701, and RA = 1\fdg 35, Dec = -26\fdg 701 respectively.

@@

The data are flagged\footnote{\url{http://sourceforge.net/p/aoflagger}} for interference \citep{off12,off10}, removing 3\% of the data and averaged in time  and frequency from the raw 0.5~s, 40~kHz to 2s, 80~kHz. These data are then calibrated to a simulation of the sky containing 2420 point--like objects selected from the MWA Commissioning Survey (MWACS; Hurley-Walker et al. 2014, submitted). MWACS was compiled using observations taken with subsets of the array during the final buildout and commissioning; it has a flux density limit of 25~mJy and a declination range of -12\arcdeg to -40\arcdeg evenly covering the field of view of the observations reported here. The objects used in calibration are selected to lie inside the 5\% contour of the primary lobe of the tile power--pattern. The calibration algorithm -- based on forward modeling software by \citet{sul12} -- computes complex gain solutions per channel per antenna averaged to two minute intervals. The solutions are fairly low signal--to--noise so we iteratively project along the antenna and frequency dimensions to capture the relatively independent passband and antenna--to--antenna variation. First, we average the channel gains over all antennas to obtain a high signal--to--noise measurement of the bandpass. After applying this single passband, we do a second round of calibration and fit second and first order polynomials for amplitude and phase respectively for each antenna. This flattens any residual variation in bandpass and removes small phase slopes due to variations in cable delay. Finally, we fit for an additional phase known to be caused by reflections in 150~m length cables. %possibly could add elaboration, if I knew exactly which version of the cal was done here. doesn't really matter

@@

We show results of delay spectrum from MWA data after deconvolution along the delay axis in Figure~\ref{fig:fhd_data}. The principle feature is the {\it foreground wedge}, whose boundaries, marked by solid white lines, increase linearly with light travel time corresponding to baseline length. Note that the wedge is not filled out uniformly. Each baseline has a different orientation and resulting response to different objects in the sky. It is this variation that we will explore below.

@@

Figure~\ref{fig:noiseless-dsm-delay-spectrum} shows the amplitude of delay spectrum, $|V_\eta(|\overline{\mathbf{x}}|,\eta)|$, simulated with only the diffuse foreground model, omitting compact sources. The top and bottom panels correspond to {\it off--zenith} and {\it zenith} pointings, respectively.
